Vragenlijst thesis evaluatie

### 2015-08-15 doc/template/chap-10.tex
vanalles veranderd

%\subsubsection{Complication}
%Since it is so difficult to protect a system or a network against APT's, researchers have been looking for effective ways to predict in advance which defence strategy might be the better one. 
There are a number of key strategies an organisation can apply to defend itself against APT's: awareness, whitelisting, system administration, network segregation, dynamic content checking and patch management. Nevertheless the combination of all these elements benefits from being complemented by other defence strategies to protect oneself against stealthy takeovers. One possible way to study the impact of stealthy takeovers and to determine practical recommendations for defenders is through game theory.\\
Game theory is gaining increasing interest as an effective technique to model and study cyber security problems. It is common to model cyber security problems as a game with two players, an attacker and a defender. There are, however, games that have more players e.g. when a third party is involved \cite{fengstealthy}. This paper focusses on a game with two players. The actions available to the attacker and the defender correspond respectively to the attacks on the system and the defensive measures that protect the system. \\

@@

This paper adapts the model presented in \cite{FlipIt} so as to take the delay for virus propagation into account.
This leads us to the following research questions:
\begin{itemize}
\item How can we incorporate the notion of delay in the game-theoretical analysis of the Flip-It game for a periodic strategy?
\item Does the resulting model allow an optimal defence strategy against an attacker? 
\end{itemize}

\subsubsection{Contributions and results}
The following contributions are made in this paper:
\begin{itemize}
\item[-] We propose an addition to the basic FlipIt model to model a scenario where the moves by the attacker will not be instantaneous. We extend the FlipIt game to a game wherein the attacker flips with a delay. The attacker only compromises the system if sufficient nodes in the network are infected. 
\item[-] The periodic case of FlipIt is modelled with a delay, resulting in optimum functions and Nash equilibria. \todo{punt herschrijven en meer nadruk leggen op het vinden van de Nash equilibria}
\item[-] Based on our results we can give practical recommendations to take measures against advanced attacks.  The modelling of FlipIt with a propagation delay will give a guideline for both the attacker and the defender on how to respectively plan attacks or network clean-ups.
\item[-] This work includes an overview of different techniques used to calculate the propagation delay of malware (depending on the network layout). \todo{en het wordt ook toegepast op de formules}
\item[-] It presents a method to calculate the speed of the propagation of a worm in a network independent of the topology of the network.
\end{itemize}

@@

The organisation of this paper is the following.  An introduction to cyber security and game theory is given in chapter \ref{Chapter1:Intro.Game.Theory} to allow the reader to become familiar with the kind of threats that are in the scope of this work and the game theoretic concepts that will be further used in the paper. In the same chapter the FlipIt framework is summarized with its most important conclusions. The chapter concludes with an overview of the related work on FlipIt and further clarifies the contributions of this paper compared to existing work. 
Chapter \ref{chapter2:FlipIt with virus propagation} first introduces the adaptations made to the original FlipIt model to model a FlipIt game with virus propagation. Subsequently formulas are derived to model a FlipIt game with a virus propagation for the specific case where players play a periodic strategy with a random phase. \\
In Chapter \ref{chapter:Nash} the formulas are further analysed to determine if there is a Nash equilibrium. 
In order to provide a clear perspective on delays, chapter \ref{chapter4: Worm propagation} gives an overview of the various methods of propagation and worm propagation models. It presents a method to calculate the speed of the propagation of a worm in a network independent of the topology of the network.
Finally chapter \ref{chapter5:conclusion} discusses the main results and provides directions for further research.
\todo{chapters kloppen nog niet}

>>>>

%\subsubsection{Complication}
%Since it is so difficult to protect a system or a network against APT's, researchers have been looking for effective ways to predict in advance which defence strategy might be the better one. 
There are a number of key strategies an organisation can apply to defend itself against APTs: awareness, whitelisting, system administration, network segregation, dynamic content checking and patch management. Nevertheless the combination of all these elements benefits from being complemented by other defence strategies to protect oneself against stealthy takeovers. One possible way to study the impact of stealthy takeovers and to determine practical recommendations for defenders is through game theory.\\
Game theory is gaining increasing interest as an effective technique to model and study cyber security problems. It is common to model cyber security problems as a game with two players, an attacker and a defender. There are, however, games that have more players e.g. when a third party is involved \cite{fengstealthy}. This paper focusses on a game with two players. The actions available to the attacker and the defender correspond respectively to the attacks on the system and the defensive measures that protect the system. \\

@@

This paper adapts the model presented in \cite{FlipIt} so as to take the delay for APT propagation into account.
This leads us to the following research questions:
\begin{itemize}
\item How can we incorporate the notion of delay in the game-theoretical analysis of the FlipIt game for a periodic strategy?
\item Does the resulting model allow an optimal defence strategy against an attacker? 
\end{itemize}

\subsubsection{Contributions and results}
The following contributions are made in this paper:
\begin{itemize}
\item[-] We propose an addition to the basic FlipIt model to model a scenario where the moves by the attacker will not be instantaneous. We extend the FlipIt game to a game wherein the attacker flips with a delay. The attacker only compromises the system if sufficient nodes in the network are infected. 
\item[-] The periodic case of FlipIt is modelled with a delay, resulting in optimum functions and Nash equilibria. \todo{punt herschrijven en meer nadruk leggen op het vinden van de Nash equilibria}
\item[-] Based on our results we can give practical recommendations to take measures against advanced attacks.  The modelling of FlipIt with a propagation delay will give a guideline for both the attacker and the defender on how to respectively plan attacks or network clean-ups.
\item[-] This work includes an overview of different propagation techniques of worms and models used to calculate the propagation delay (depending on the network layout). It also presents a method to calculate the speed of the propagation of a worm in a propagation independent of the topology of the network.
\item[-] The Google Page Rank matrix is introduced as a proof of concept to calculate the importance of every node in the network.
\end{itemize}

@@

The organisation of this paper is the following.  An introduction to cyber security and game theory is given in chapter \ref{Chapter1:Intro.Game.Theory} to allow the reader to become familiar with the kind of threats that are in the scope of this work and the game theoretic concepts that will be further used in the paper. In the same chapter the FlipIt framework is summarized with its most important conclusions. The chapter concludes with an overview of the related work on FlipIt and further clarifies the contributions of this paper compared to existing work. 
Chapter \ref{chapter2:FlipIt with virus propagation} first introduces the adaptations made to the original FlipIt game to model a FlipIt game with a propagation delay. Subsequently formulas are derived to model a FlipIt game with a propagation delay for the specific case where players play a periodic strategy with a random phase. \\
In Chapter \ref{chapter:Nash} the formulas are further analysed to determine if there is a Nash equilibrium. 
In order to provide a clear perspective on delays, chapter \ref{chapter4: Worm propagation} gives an overview of the various methods of propagation and worm propagation models. It presents a method to calculate the speed of the propagation of a worm in a network independent of the topology of the network.
Finally chapter \ref{chapter5:conclusion} discusses the main results and provides directions for further research.

### 2015-08-15 doc/template/chap-5.tex
vanalles veranderd

\chapter{General context}
\label{Chapter1:Intro.Game.Theory}
%\documentclass[10pt]{article}
%\begin{document}

@@

This chapter provides the reader with an introduction to the general context of the work presented in this paper.  Section 2.1 introduces the reader into the basic concepts of cyber security and the kind of cyber security threats that are in scope of this work. Section 2.2 then introduces the reader into the main principles of game theory. Subsequently, section 2.3 introduces the reader to the FlipIt game: this game will be used to model cyber security attacks of a periodic nature, including a delay.  Finally, section 2.4 gives an overview of the related work, and how the research presented in this paper compares to existing results.
%- Remark: to the reader weg
\section{What Is Security?}

@@

 %The purpose of security is to give certainty that data will not be removed without authorization (Confidentiality) that the data is always accessible (Availability), and that data can not be read or altered by someone who does not have the authorization (Integrity). 
 ~~\\
 These key attributes are also known as the CIA triad. They are the fundamental security objectives for securing data, information and computing services. \\
 \\
Information security can be divided into two main subcategory. One of them is cyber security, also known as computer security. This is the process of applying security tools to ensure confidentiality, integrity, and availability of data. It is an attempt to protect websites, servers, data, applications, operating systems and all assets that need protection in a computer system.  Some of these tools may include detection, identification or removal tools. A detection tool will determine if an infection has taken place and will trace the threat. An identification tool will try to identify the threat to be able to know how to remove it. A removal tool will remove the threat from the system (once it has been identified) so that it cannot spread any further. \\
The other category is network or internet security. This sub category of information security protects data during transmission. While cyber security and network security address different aspects, they partially overlap as well. For example, a virus can be physically dropped on a computer network using a USB stick, but it can also arrive over the internet. Either way internal computer security measurements have to be taken to recover from the virus. In this paper we focus on scenarios where a network has to be defended against attacks to ensure confidentiality, integrity and availability of data. The work presented in this paper therefore belongs to the domain of cyber security.

@@

The most noteworthy and biggest group of threats to computer systems is malware. 
This is a piece of malicious software that is designed to penetrate unprotected or vulnerable systems or computers, with the intent to retrieve sensitive information, destroy data, or compromise the confidentiality, integrity or availability of the data or applications of the victim. A security report of 2014 \cite{SurveyKaspersky} reveals that 61\% of the attacks on companies are caused by malware. For this reason this section will examine the categories of malware threats. Different types of malware exist: viruses, worms, flooders, rootkits, bots, spyware, adware and many more. This broad range of different types can be classified into two main categories. The first one based on the propagation method that is used and the second one based on the payload or the variety of actions that the malware performs \cite{stallings2007network}. Propagation methods include viruses, worms and trojans. Payload includes e.g. flooders, rootkits, bots, spyware and adware. This paper focuses on the category of propagation methods and not on the actions that the malware performs. A brief explanation of the three main propagation methods of malware is given below.

>>>>

\chapter{General Context}
\label{Chapter1:Intro.Game.Theory}
%\documentclass[10pt]{article}
%\begin{document}

@@

This chapter provides the reader with an introduction to the general context of the work presented in this paper.  First, section 2.1 introduces the reader into the basic concepts of cyber security and the kind of cyber security threats that are in scope of this work. Section 2.2 then introduces the reader into the main principles of game theory. Subsequently, section 2.3 introduces the reader to the FlipIt game: this game will be used to model cyber security attacks of a periodic nature, including a delay.  Finally, section 2.4 gives an overview of the related work, and how the research presented in this paper compares to existing results.
%- Remark: to the reader weg
\section{What Is Security?}

@@

 %The purpose of security is to give certainty that data will not be removed without authorization (Confidentiality) that the data is always accessible (Availability), and that data can not be read or altered by someone who does not have the authorization (Integrity). 
 ~~\\
 These key attributes are also known as the CIA triad. They are the fundamental security objectives for securing data, information and computing services. \\
 \\
Information security can be divided into two main subcategories. One of them is cyber security, also known as computer security. This is the process of applying security tools to ensure confidentiality, integrity, and availability of data. It is an attempt to protect websites, servers, data, applications, operating systems and all assets that need protection in a computer system.  Some of these tools may include detection, identification or removal tools. A detection tool will determine if an infection has taken place and will trace the threat. An identification tool will try to identify the threat to be able to know how to remove it. A removal tool will remove the threat from the system (once it has been identified) so that it cannot spread any further. \\
The other category is network or internet security. This subcategory of information security protects data during transmission. While cyber security and network security address different aspects, they partially overlap as well. For example, a virus can be physically dropped on a computer network using a USB stick, but it can also arrive over the internet. Either way internal computer security measurements have to be taken to recover from the virus. In this paper we focus on scenarios where a network has to be defended against attacks to ensure confidentiality, integrity and availability of data. The work presented in this paper therefore belongs to the domain of cyber security.

@@

The most noteworthy and biggest group of threats to computer systems is malware. 
This is a piece of malicious software that is designed to penetrate unprotected or vulnerable systems or computers, with the intent to retrieve sensitive information, destroy data, or compromise the confidentiality, integrity or availability of the data or applications of the victim. A security report of Kaspersky in 2014 \cite{SurveyKaspersky} reveals that 61\% of the attacks on companies are caused by malware. For this reason this section will examine the categories of malware threats. Different types of malware exist: viruses, worms, flooders, rootkits, bots, spyware, adware and many more. This broad range of different types can be classified into two main categories. The first one based on the propagation method that is used and the second one based on the payload or the variety of actions that the malware performs \cite{stallings2007network}. Propagation methods include viruses, worms and trojans. Payload includes e.g. flooders, rootkits, bots, spyware and adware. This paper focuses on the category of propagation methods and not on the actions that the malware performs. A brief explanation of the three main propagation methods of malware is given below.
